---
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# F2Adonis2

F2Adonis2 is a tool for multivariate hypothesis testing, developed to support clinical metabolomics and high-dimensional omics research. It addresses the multivariate Behrens-Fisher Problem in dissimilarity-based analyses (such as PERMANOVA), providing a statistically sound approach for datasets where group dispersions are heterogeneous. It is only available on GitHub.

## Motivation

In high-dimensional ecological and metabolomics studies ($P \gg N$), standard PERMANOVA ($F_1$) methods often yield inflated Type I error rates when group dispersions are unequal (heteroscedasticity) and group sizes are unbalanced.

This package implements the methodological solutions proposed by **Anderson et al. (2017)**. Rather than pooling variance across groups (which leads to bias), the $F_2$ statistic uses a weighted combination of group-specific dispersion estimates to maintain level accuracy under arbitrary heteroscedastic conditions.


## Installation

You must install the `remotes` package first.

```{r, eval = FALSE}
install.packages("remotes", repos = "[https://cloud.r-project.org](https://cloud.r-project.org)")
```

Then, use `install_github` to install this package.

```{r, eval = FALSE}
remotes::install_github("LakviruPerera/F2Adonis2")
```



## Quick Start

`F2Adonis2` acts as a drop-in replacement for `vegan::adonis2`. It expects a distance matrix (typically from `vegdist`) and follows the standard R formula interface.


```{r example, eval=FALSE}

library(F2Adonis2)
library(vegan)

# Calculate dissimilarity matrix
dist_sub <- vegdist(sub_metabolites, method = "euclidean")


# Run the robust F2 test
fit_f2 <- F2_adonis2(dist_sub ~ Group, 

                     data = sub_metadata, 

                     permutations = 999)

# View Results
print(fit_f2)

```


## References

Anderson, M.J., Walsh, D.C.I., Robert Clarke, K., Gorley, R.N. and Guerra-Castro, E. (2017), Some solutions to the multivariate Behrens–Fisher problem for dissimilarity-based analyses. Aust. N. Z. J. Stat., 59: 57-79. https://doi.org/10.1111/anzs.12176

Oksanen J, Simpson G, Blanchet F, Kindt R, Legendre P, Minchin P, O'Hara R, Solymos P, Stevens M, Szoecs E, Wagner H, Barbour M, Bedward M, Bolker B, Borcard D, Borman T, Carvalho G, Chirico M, De Caceres M, Durand S, Evangelista H, FitzJohn R, Friendly M, Furneaux B, Hannigan G, Hill M, Lahti L, Martino C, McGlinn D, Ouellette M, Ribeiro Cunha E, Smith T, Stier A, Ter Braak C, Weedon J (2026). vegan: Community Ecology Package. R package version 2.8-0, https://vegandevs.github.io/vegan/.

