---
title: "PKD_Subsample_Study"
author: "Lakviru Perera"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PKD_Subsample_Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

This vignette demonstrates the divergence between standard PERMANOVA ($F_1$)
and the modified F2 statistic under a deliberately unbalanced, dispersion-
heterogeneous subsample of the PKD urinary metabolomics dataset (10 Ctrl / 20
PKD), and cross-checks the result using dispersion diagnostics, ordination,
and component-level decomposition.

To keep this vignette fast to build, all permutation tests below use
`permutations = 99` rather than the `permutations = 999` used to generate the
results reported in the accompanying JOSS paper. Results here will therefore
approximate, but not exactly reproduce, the paper's reported statistics and
p-values. A code snippet reproducing the paper's exact settings (999
permutations) is included at the end of this vignette for reference.

```{r}
#| label: fig-pkd-subsample
#| fig-cap: "Check on a subsampled PKD dataset (10 Ctrl / 20 PKD): dispersion diagnostics, F1 vs. F2 inference, PCA ordination, and component-level decomposition."

library(dplyr)
library(vegan)
library(F2Adonis2)
library(kableExtra)
library(ggplot2)
library(combinat)

# Load package data objects explicitly
data("pkd_metadata", package = "F2Adonis2")
data("pkd_metabolites", package = "F2Adonis2")

set.seed(527)

# ==============================================================================
# 1. Sub-sample the PKD dataset (10 Ctrl / 20 PKD)
# ==============================================================================
sub_metadata <- bind_rows(
  pkd_metadata %>% filter(Group == "Ctrl") %>% slice_sample(n = 10),
  pkd_metadata %>% filter(Group == "PKD") %>% slice_sample(n = 20)
)
sub_metabolites <- pkd_metabolites[rownames(sub_metadata), ]
dist_sub <- vegdist(sub_metabolites, method = "euclidean")

# ==============================================================================
# 2. Dispersion diagnostics (betadisper)
# ==============================================================================
mod_sub <- betadisper(dist_sub, sub_metadata$Group)
anova_disp <- permutest(mod_sub, permutations = 99)
print(anova_disp)

# Spatial hull plot
plot(mod_sub, hull = TRUE, ellipse = FALSE,
     main = "Spatial Dispersion: Sub-sample (10 Ctrl / 20 PKD)")

# Violin + boxplot + jittered points of distance-to-centroid
distances_df <- data.frame(
  distance = mod_sub$distances,
  group = mod_sub$group
)

ggplot(distances_df, aes(x = group, y = distance, fill = group)) +
  geom_violin(alpha = 0.2, trim = TRUE, color = NA) +
  geom_boxplot(width = 0.2, alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.1, size = 1.5, alpha = 0.5) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank()
  ) +
  labs(
    title = "Dispersion Check: Sub-sampled Data",
    x = "Group",
    y = "Distance to Centroid"
  )

# ==============================================================================
# 3. F1 vs. F2 inference on the subsample
# ==============================================================================
pkd_fit_f1_sub <- adonis2(dist_sub ~ Group, data = sub_metadata,
                          method = "euclidean", permutations = 99)

pkd_fit_f2_sub <- F2_adonis2(dist_sub ~ Group, data = sub_metadata,
                             method = "euclidean", permutations = 99)

f1_p_sub <- pkd_fit_f1_sub["Model", "Pr(>F)"]
f2_p_sub <- pkd_fit_f2_sub["Model", "Pr(>F2)"]

sub_results <- data.frame(
  Sample_Size = "10 Ctrl / 20 PKD",
  F1_Pval = f1_p_sub,
  F2_Pval = f2_p_sub,
  Verdict = ifelse(f2_p_sub > 0.05 & f1_p_sub < 0.05, "Conflicted", "Consistent")
)

sub_results %>%
  mutate(
    F1_Pval = sprintf("%.4f%s", F1_Pval, ifelse(F1_Pval < 0.05, "*", "")),
    F2_Pval = sprintf("%.4f", F2_Pval)
  ) %>%
  kable(
    col.names = c(
      "Subsample Configuration",
      "Standard F1 ($p$)",
      "F2 ($p$)",
      "Verdict"
    ),
    escape = FALSE,
    caption = "Check: Subsampled PKD Analysis"
  ) %>%
  kable_styling(full_width = FALSE)

print(pkd_fit_f1_sub)
print(pkd_fit_f2_sub)

# F1 permutation null distribution
psub_f1 <- permustats(pkd_fit_f1_sub)
f1_plot_data <- data.frame(F_vals = as.numeric(psub_f1$permutations))
f1_obs_stat <- pkd_fit_f1_sub["Model", "F"]

ggplot(f1_plot_data, aes(x = F_vals)) +
  geom_histogram(bins = 30, fill = "darkgoldenrod", color = "white", alpha = 0.7) +
  geom_vline(xintercept = f1_obs_stat, color = "red", linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = f1_obs_stat, y = Inf, label = paste("Observed F =", round(f1_obs_stat, 3)),
           vjust = 2, hjust = -0.1, color = "red", fontface = "bold") +
  theme_minimal() +
  labs(
    title = "F1 Null Distribution: PKD Subsample",
    x = "Permuted Pseudo-F Statistic",
    y = "Frequency"
  )

# ==============================================================================
# 4. PCA ordination and component-level decomposition
# ==============================================================================
pca_fit <- prcomp(sub_metabolites, scale. = TRUE)
pca_scores <- pca_fit$x[, 1:5]
comp_data <- data.frame(pca_scores, Group = sub_metadata$Group)
var_exp <- (pca_fit$sdev^2 / sum(pca_fit$sdev^2)) * 100

# PCA scores plot (PC1 vs PC2)
ggplot(comp_data, aes(x = PC1, y = PC2, color = Group, fill = Group)) +
  geom_point(size = 3, alpha = 0.6) +
  stat_ellipse(geom = "polygon", alpha = 0.2) +
  theme_minimal() +
  labs(title = "PCA Scores: Subsampled Data",
       subtitle = "Visualizing the overlap of group centroids",
       x = paste0("PC1 (", round(var_exp[1], 1), "%)"),
       y = paste0("PC2 (", round(var_exp[2], 1), "%)"))

# Univariate t-tests on the top 5 PCs
univariate_tests <- lapply(1:5, function(i) {
  test <- t.test(comp_data[[i]] ~ comp_data$Group)
  data.frame(Component = paste0("PC", i), P_Value = test$p.value)
})

do.call(rbind, univariate_tests) %>%
  mutate(P_Value = sprintf("%.4f%s", P_Value, ifelse(P_Value < 0.05, "*", ""))) %>%
  kable(caption = "Decomposition: Univariate Significance of Components",
        booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position")

# Pairwise F2 sensitivity analysis across all 2D PC planes
pc_pairs <- combn(1:5, 2)

pairwise_results <- lapply(1:ncol(pc_pairs), function(i) {
  pair_idx <- pc_pairs[, i]
  dist_pair <- dist(pca_scores[, pair_idx], method = "euclidean")
  
  fit <- F2_adonis2(dist_pair ~ Group, data = sub_metadata,
                    permutations = 99, bootstrap = FALSE, plot.dist = FALSE)
  
  data.frame(
    Pair = paste0("PC", pair_idx[1], " vs PC", pair_idx[2]),
    F2_Stat = fit["Model", "F2"],
    P_Value = fit["Model", "Pr(>F2)"]
  )
})

do.call(rbind, pairwise_results) %>%
  mutate(P_Value = sprintf("%.4f%s", P_Value, ifelse(P_Value < 0.05, "*", ""))) %>%
  kable(caption = "Pairwise Sensitivity Analysis: F2 on 2D PC Planes",
        booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position")
```

## Reproducing the paper's exact settings

The chunk below reproduces the same analysis using `permutations = 999`, the
setting used to generate the results reported in the accompanying JOSS paper.
It is not executed automatically when this vignette is built, since running
999 permutations across the twelve model fits above is unnecessarily slow for
routine package documentation. Copy this code into an R session to run it
directly.

```{r pkd-subsample-full, eval = FALSE}
pkd_fit_f1_sub <- adonis2(dist_sub ~ Group, data = sub_metadata,
                          method = "euclidean", permutations = 999)

pkd_fit_f2_sub <- F2_adonis2(dist_sub ~ Group, data = sub_metadata,
                             method = "euclidean", permutations = 999)

pairwise_results_full <- lapply(1:ncol(pc_pairs), function(i) {
  pair_idx <- pc_pairs[, i]
  dist_pair <- dist(pca_scores[, pair_idx], method = "euclidean")

  fit <- F2_adonis2(dist_pair ~ Group, data = sub_metadata,
                    permutations = 999, bootstrap = FALSE, plot.dist = FALSE)

  data.frame(
    Pair = paste0("PC", pair_idx[1], " vs PC", pair_idx[2]),
    F2_Stat = fit["Model", "F2"],
    P_Value = fit["Model", "Pr(>F2)"]
  )
})

do.call(rbind, pairwise_results_full)
```
