---
title: "Introduction_to_F2Adonis2"
author: "Lakviru Perera"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction_to_F2Adonis2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(

  collapse = TRUE,

  comment = "#>",

  fig.width = 7,

  fig.height = 5

)
```


```{r setup}
library(ggplot2)
library(tidyverse)
library(F2Adonis2)
```


#Summary#

F2Adonis2 is an R package designed for robust multivariate analysis of variance using distance matrices. It implements the $F_2$ statistic as proposed by Anderson et al. (2017) to provide a solution to the multivariate Behrens-Fisher problem (BFP). The BFP arises when researchers need to test the null hypothesis of equality of mean vectors (centroids) across two or more populations while their dispersion matrices (variances) differ.

#Statement of Need#

In multivariate ecological and biological studies, researchers often rely on distance-based permutational methods like PERMANOVA (the $F_1$ statistic) to detect differences among groups. However, standard PERMANOVA is strictly valid only when group dispersions are homogeneous. As demonstrated in Anderson et al. (2017), the $F_1$ test is not robust to heterogeneity of dispersions in the space of the chosen dissimilarity measure, particularly for unbalanced designs:

Inflated Type I Error: When smaller groups have larger spreads, the $F_1$ test becomes too liberal, falsely identifying differences in centroids.

Loss of Power: When larger groups have larger spreads, the $F_1$ test becomes too conservative.

F2Adonis2 provides a mathematically grounded solution by modifying the PERMANOVA test statistic, ensuring valid inference even when groups exhibit heteroscedasticity in non-Euclidean space.

#Methodology: The Robust $F_2$ Statistic#

The $F_2$ statistic addresses the BFP by replacing the pooled residual error term in the denominator with a weighted average of group-specific mean-squared dissimilarities ($V_i$):

$$F_2 = \frac{SS_A / (g - 1)}{\sum_{i=1}^g (1 - n_i/N)V_i}$$

Where $SS_A$ is the among-group sum of squares, $g$ is the number of groups, $n_i$ is the sample size of group $i$, $N$ is the total sample size, and $V_i$ is the group-specific variance calculated in dissimilarity space.

Resampling Methods

For metric distances, the package supports standard permutation. However, for non-metric or semi-metric distances (like Jaccard or Bray-Curtis), the package implements matrix-space residual bootstrapping. This procedure uses a residualized G matrix (Gower's centered matrix) to preserve group-specific dispersions while strictly maintaining the null hypothesis of equal centroids.

#Case Study: Norway Benthic Macrofauna#

To demonstrate the robustness of the $F_2$ statistic, we replicate the analysis of the Norway benthic macrofauna dataset (Ekofisk oil field). This dataset consists of presence-absence data for 809 species across 101 sites, categorized into 5 Areas with unbalanced sample sizes.

1. Load Data and Libraries

```{r}

library(F2Adonis2)

library(vegan)



# Load package datasets

data(norway_pa)

data(norway_metadata)



# Unbalanced design overview

table(norway_metadata$Area)
```

2. Visualizing Dispersion Heterogeneity

Before running the test, we can confirm the presence of the Behrens-Fisher problem by visualizing the unequal group spreads using vegan::betadisper.

```{r}
dist_jaccard <- vegdist(norway_pa, method = "jaccard")

mod_disp <- betadisper(dist_jaccard, norway_metadata$Area)

# Plotting dispersions
plot(mod_disp, main = "Multivariate Dispersions (Norway Data)")


library(yarrr)

# Extract data into a frame for plotting
disp_data <- data.frame(
  Distance = mod_disp$distances,
  Area = mod_disp$group
)

# Enhanced Pirate Plot
pirateplot(formula = Distance ~ Area,
           data = disp_data,
           main = "Multivariate Dispersions (Norway Data)",
           xlab = "Geographic Area",
           ylab = "Distance to Centroid",
           theme = 1,          # Clean theme
           pal = "basel",      # Professional palette
           point.o = 0.9,      # Transparency of raw points
           avg.line.o = 0.8,   # High visibility for mean
           bean.f.o = 0.2,     # Light density bean
           inf.method = "hdi") # High-density interval for inference
```


3. Global Robust PERMANOVA

We apply the F2_adonis2 function. We enable bootstrap = TRUE and bias.adjust = TRUE to account for the non-metric nature of the Jaccard distance and to correct for potential variance overestimation in small samples.


```{r}
set.seed(123)

# Conventional perMANOVA pseudo-F:

fit_results1 <- adonis2(

  formula = norway_pa ~ Area,

  data = norway_metadata,

  method = "jaccard",

  permutations = 999

)

fit_results1


# Running the robust F2 test

fit_results2 <- F2_adonis2(

  formula = norway_pa ~ Area,

  data = norway_metadata,

  method = "jaccard",

  permutations = 999,

  bootstrap = TRUE,

  bias.adjust = TRUE

)



fit_results2
```

* Comparison of permutation distributions:


```{r}
attributes(fit_results1)

attributes(fit_results1)$F.perm %>% as.data.frame() %>% ggplot(aes(x = V1)) + 
  geom_histogram() + 
  geom_vline(xintercept = fit_results1$F[1])

#attributes(fit_results1)$F.perm %>% as.data.frame() %>% ggplot(aes(x = V1)) + 
#  geom_histogram() + 
#  geom_vline(xintercept = fit_results1$F[1])


```


#Univariate Comparison#

The test statistic provides a generalization of Welch's F-test. The following code compares the two results and the more conventional approaches in two situations, one with no difference in the means but different variances and one with different means and different variances.

```{r}
# # Simulate three unbalanced groups, same means, different variances:
# set.seed(93939)
# Xf <- factor(c(rep("Lev1", 15), rep("Lev2", 10), rep("Lev3", 5)))
# 
# y1 <- rnorm(n = 30, mean = rep(0, 30), sd = 20*as.numeric(Xf))
# 
# df1 <- data.frame(y1, Xf)
# 
# df1 %>% ggplot(aes(x = Xf, y = y1)) +
#   geom_violin() +
#   geom_point()
# 
# anova(lm(y1 ~ Xf, data = df1))
# 
# ####################################
# # Welch test
# library(onewaytests)
# 
# welch.test(y1 ~ Xf, data = df1)
# 
# # Try permutations with Welch's F-test as well:
# 
# # Custom function to get Welch's F statistic
# get_welch_f <- function(y, groups) {
#   oneway.test(y ~ groups)$statistic
# }
# 
# # Permutation procedure for Welch's F
# set.seed(123)
# n_perm <- 999
# welch_obs <- get_welch_f(y1, Xf)
# welch_perms <- replicate(n_perm, {
#   get_welch_f(y1, sample(Xf))
# })
# 
# p_welch_perm <- (sum(welch_perms >= welch_obs) + 1) / (n_perm + 1)
# cat("Permutational Welch's F p-value:", p_welch_perm)
# 
# #####################################
# # perMANOVA for Univariate F-test
# 
# dist1 <- dist(y1, method = "euclidean")
# set.seed(123)
# adonis2(dist1 ~ Xf, data = df1)
# 
# # F2 for Univariate F-test
# set.seed(123)
# F2_adonis2(
#   formula = y1 ~ Xf,
#   data = df1,
#   method = "euclidean",
#   sqrt.dist = F,
#   permutations = 999,
#   bootstrap = TRUE,
#   bias.adjust = TRUE
# )
# set.seed(123)
# F2_adonis2(
#   formula = dist1 ~ Xf,
#   data = df1,
#   method = "euclidean",
#   sqrt.dist = F,
#   permutations = 999,
#   bootstrap = TRUE,
#   bias.adjust = TRUE
# )

```


```{r}
################################################################################
# F2Adonis2 Vignette: Univariate Validation and Welch's F Comparison
################################################################################

library(ggplot2)
library(tidyverse)
library(vegan)
library(F2Adonis2)
library(onewaytests)

# ==============================================================================
# 1. Simulate three unbalanced groups, same means, different variances:
# ==============================================================================
set.seed(93939)
Xf <- factor(c(rep("Lev1", 15), rep("Lev2", 10), rep("Lev3", 5)))
y1 <- rnorm(n = 30, mean = rep(0, 30), sd = 20 * as.numeric(Xf))
df1 <- data.frame(y1, Xf)

# Visualize the heterogeneity (The Behrens-Fisher Problem)
df1 %>% ggplot(aes(x = Xf, y = y1, fill = Xf)) +
  geom_violin(alpha = 0.3) +
  geom_point() +
  theme_minimal() +
  labs(title = "Simulation: Same Means, Heterogeneous Variance")

# Standard ANOVA - significant (p=0.031) because it is biased by the variance
anova(lm(y1 ~ Xf, data = df1))

# ==============================================================================
# 2. Welch's F-test and Permutational Null
# ==============================================================================
welch_res <- oneway.test(y1 ~ Xf, data = df1)
print(welch_res)

get_welch_f <- function(y, groups) {
  oneway.test(y ~ groups)$statistic
}

# Generate Permutational Null Distribution for Welch's F
set.seed(123)
n_perm <- 999
welch_obs <- get_welch_f(y1, Xf)
welch_perms <- replicate(n_perm, {
  get_welch_f(y1, sample(Xf))
})

p_welch_perm <- (sum(welch_perms >= welch_obs) + 1) / (n_perm + 1)
cat("Permutational Welch's F p-value:", p_welch_perm, "\n")

# ==============================================================================
# 3. F2 (F2Adonis2) Univariate Implementation & Manual Null Capture
# ==============================================================================
# Get Observed F2 - Using permutations = 1 to avoid subscript errors
set.seed(123)
fit_f2 <- F2_adonis2(y1 ~ Xf, data = df1, method = "euclidean", permutations = 1)
f2_obs <- fit_f2["Model", "F2"]

# FIX: Manually capture the F2 null distribution
# We shuffle the data frame manually and run F2 with permutations = 1
cat("Generating F2 null distribution...\n")
set.seed(123)
f2_perms <- numeric(n_perm)

for(i in 1:n_perm) {
  # Shuffle labels manually
  shuffled_df <- df1
  shuffled_df$Xf <- sample(df1$Xf)
  
  # Run and extract the statistic for this shuffle
  tmp <- F2_adonis2(y1 ~ Xf, data = shuffled_df, method = "euclidean", permutations = 1)
  f2_perms[i] <- tmp["Model", "F2"]
}

# ==============================================================================
# 4. Compare Percentiles and Distributions
# ==============================================================================
# Observed Percentile (Quantile): What percentage of the null is below the observed?
perc_welch <- mean(welch_perms <= welch_obs, na.rm = TRUE)
perc_f2    <- mean(f2_perms <= f2_obs, na.rm = TRUE)

cat("\n--- Comparison of Observed Quantiles ---\n")
cat(sprintf("Welch's F Observed Percentile: %.4f\n", perc_welch))
cat(sprintf("F2 Statistic Observed Percentile: %.4f\n", perc_f2))

# Visualize comparison of null distributions
null_df <- data.frame(
  Value = c(welch_perms, f2_perms),
  Test = c(rep("Welch F", length(welch_perms)), rep("F2 Stat", length(f2_perms)))
)

ggplot(null_df, aes(x = Value, fill = Test)) +
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = welch_obs, color = "blue", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = f2_obs, color = "red", linetype = "dashed", linewidth = 1) +
  theme_minimal() +
  labs(title = "Permutational Distributions: Welch F vs F2",
       subtitle = "Dashed lines represent observed statistics",
       x = "Statistic Value", y = "Density")

```


```{r}
# ==============================================================================
# 5. High-Dimensional Case Study (PKD Metabolomics)
# ==============================================================================
# Demonstrating scaling to 932 variables where Welch's F is unavailable.
library(dplyr)
library(F2Adonis2)

set.seed(123)

cat("\n--- Loading and Preprocessing PKD Data ---\n")

# Read raw data
pkd_raw <- read.csv("PKD Urine Metabolomics LCMS Data_110221_2groups.csv", 
                    check.names = FALSE, stringsAsFactors = FALSE)

# Extract Metadata and Features
group_labels <- as.character(pkd_raw[1, -1])
group_labels <- factor(group_labels)
sample_names <- colnames(pkd_raw)[-1]
metabolite_names <- pkd_raw[-1, 1]

# Format Matrix (Samples as rows, Metabolites as columns)
Y_data <- as.matrix(pkd_raw[-1, -1])
mode(Y_data) <- "numeric"
Y_data <- t(Y_data)
rownames(Y_data) <- sample_names
colnames(Y_data) <- metabolite_names

# Missing Value Handling (>50% threshold)
missing_pct <- colMeans(is.na(Y_data))
Y_data <- Y_data[, missing_pct <= 0.5]

# Imputation (method='min', replaced with 1/5 of min positive value)
for(i in 1:ncol(Y_data)) {
  col_data <- Y_data[, i]
  if(any(is.na(col_data))) {
    min_positive <- min(col_data[col_data > 0], na.rm = TRUE)
    Y_data[is.na(col_data), i] <- min_positive / 5
  }
}

# Normalization (GroupPQN using Pool samples)
pool_idx <- grepl("Pool", rownames(Y_data), ignore.case = TRUE)
if(any(pool_idx)) {
  pool_data <- Y_data[pool_idx, , drop = FALSE]
  sample_data <- Y_data[!pool_idx, , drop = FALSE]
  sample_groups <- group_labels[!pool_idx]
  
  ref_spectrum <- apply(pool_data, 2, median, na.rm = TRUE)
  
  normalize_pqn <- function(sample_row, reference) {
    quotients <- sample_row / reference
    scaling_factor <- median(quotients[is.finite(quotients) & quotients > 0], na.rm = TRUE)
    return(sample_row / scaling_factor)
  }
  
  Y_normalized <- t(apply(sample_data, 1, normalize_pqn, reference = ref_spectrum))
  Y_data <- Y_normalized
  group_labels <- factor(sample_groups)
}

# Log10 Transformation and Auto-scaling
Y_data <- log10(Y_data)
AutoNorm <- function(x){(x - mean(x))/sd(x, na.rm=T)}
Y_pkd_final <- apply(Y_data, 2, AutoNorm)

cat(sprintf("Final Processed Data: %d samples x %d metabolites\n", 
            nrow(Y_pkd_final), ncol(Y_pkd_final)))

# ==============================================================================
# 2. PERFORM ROBUST F2 ANALYSIS
# ==============================================================================
cat("\n--- Running Robust F2 PERMANOVA ---\n")

# Prepare Metadata Frame
pkd_meta <- data.frame(Group = group_labels)

# Run F2 with 999 permutations and Bias-Adjusted Bootstrapping
# Using Euclidean distance as data is log-transformed and scaled
pkd_f2_results <- F2_adonis2(
  formula = Y_pkd_final ~ Group,
  data = pkd_meta,
  method = "euclidean",
  permutations = 999,
  bootstrap = TRUE,
  bias.adjust = TRUE
)

print(pkd_f2_results)
```


#Implementation Details#

The F2_adonis2 function is designed to be a "drop-in" replacement for vegan::adonis2, supporting standard R formulas and distance calculations via vegdist. Key features include:

Internal Consistency: Implements the $V_i$ estimator directly from the distance matrix to handle high-dimensional compositional data.

Flexible Bootstrapping: Offers both standard permutation and separate-sample residual bootstrapping.

Bias Correction: Includes an empirical bias correction for bootstrap variances as derived in Anderson (2017).

#References#

Anderson, M.J., Walsh, D.C.I., Robert Clarke, K., Gorley, R.N. and Guerra-Castro, E. (2017), Some solutions to the multivariate Behrens–Fisher problem for dissimilarity-based analyses. Aust. N. Z. J. Stat., 59: 57-79. https://doi.org/10.1111/anzs.12176

Dag, O., Dolgun, A., Konar, N.M. (2018). onewaytests: An R Package for One-Way Tests in Independent Groups Designs. The R Journal, 10:1, 175-199.

Oksanen J, Simpson G, Blanchet F, Kindt R, Legendre P, Minchin P, O'Hara R, Solymos P, Stevens M, Szoecs E, Wagner H, Barbour M, Bedward M, Bolker B, Borcard D, Borman T, Carvalho G, Chirico M, De Caceres M, Durand S, Evangelista H, FitzJohn R, Friendly M, Furneaux B, Hannigan G, Hill M, Lahti L, Martino C, McGlinn D, Ouellette M, Ribeiro Cunha E, Smith T, Stier A, Ter Braak C, Weedon J (2026). vegan: Community Ecology Package. R package version 2.8-0, https://vegandevs.github.io/vegan/.

Welch, B. L.(1951). On the Comparison of Several Mean Values: An Alternative Approach. Biometrika, 38, 330-336.

#run devtools::build_vignettes()

